Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-05-30 pkg/paper/Paper/draft.Rnw
Included computations for EU in results-adjusted.Rnw and modified draft.Rnw accordingly.
Note: To work properly, the latest version of glogis needs to be installed
because the columns in HICP and hicps were reordered.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@109 b679b2f7-6539-45d0-a605-5ddcb976cde9

## additional moment method
moments <- function(object, ...) UseMethod("moments")
moments.glogisfit <- function(object, ...) object$moments
moments.breakpoints.glogisfit <- function(object, breaks = NULL, ...)
  t(sapply(refit(object, breaks = breaks), "[[", "moments"))
@

@@

<<analysis-results>>=
## run analysis (for seasonally adjusted data)
root_dir <- do.call("file.path", as.list(head(strsplit(getwd(), .Platform$file.sep)[[1]], -1)))
if(!file.exists(file.path(root_dir, "Analysis", "Austria.rda"))) {
setwd(file.path(root_dir, "Analysis"))
Sweave("results1s.Rnw")
setwd(file.path(root_dir, "Paper"))
}

## query result file names
fnam <- Sys.glob(file.path("..", "Analysis", "*.rda"))
fnam_omit <- c(grep("2", fnam, fixed = TRUE), grep("glmean", fnam, fixed = TRUE))
if(length(fnam_omit) > 0) fnam <- fnam[-fnam_omit]

## set up named results list
results <- vector(length = length(fnam), mode = "list")
names(results) <- sapply(strsplit(sapply(strsplit(fnam, "/", fixed = TRUE), tail, 1), ".", fixed = TRUE), head, 1)

## Exchange Rate Mechanism information
## ERM~II, ERM~I (part of european monetary system expired in 1999-01-01 gave way to ERM II
## if never entered ""right?
## http://web.archive.org/web/20080101022356/http://lexikon.meyers.de/meyers/Europ%C3%A4isches_W%C3%A4hrungssystem
erm <- yearmon(c(
  Austria = 1995,
  Belgium = 1979 + 2/12,
  CzechRepubic = NA, 
  Denmark = 1979 + 2/12,
  Estonia = NA,
  Finland = 1996 + 9/12,
  France = 1979 + 2/12,
  Germany = 1979 + 2/12,
  Greece = 1998 + 2/12,
  Hungary = NA,
  Ireland = 1979 + 2/12,
  Italy = 1979 + 2/12,
  Luxembourg = 1979 + 2/12,
  Netherlands = 1979 + 2/12,
  Poland = NA,
  Portugal = 1992 + 3/12,
  Slovakia = NA,
  Slovenia = NA,
  Spain = 1986 + 5/12,
  Sweden = NA ,
  UK = NA
))

erm2 <- yearmon(c(
  Austria = NA,
  Belgium = NA,
  CzechRepubic = NA, 
  Denmark = 1999,
  Estonia = 2004 + 5/12,
  Finland = NA,
  France = NA,
  Germany = NA,
  Greece = 1999,
  Hungary = NA,
  Ireland = NA,
  Italy = NA,
  Luxembourg = NA,
  Netherlands = NA,
  Poland = NA,
  Portugal = NA,
  Slovakia = 2005 + 10/12,
  Slovenia = 2004 + 5/12,
  Spain = NA,
  Sweden = NA ,
  UK = NA
))

eur <- yearmon(c(
  Austria = 1999,
  Belgium = 1999,
  CzechRepubic = NA, 
  Denmark = NA,
  Estonia = 2011,
  Finland = 1999,
  France = 1999,
  Germany = 1999,
  Greece = 2001,
  Hungary = NA,
  Ireland = 1999,
  Italy = 1999,
  Luxembourg = 1999,
  Netherlands = 1999,
  Poland = NA,
  Portugal = 1999,
  Slovakia = 2009,
  Slovenia = 2007,
  Spain = 1999,
  Sweden = NA ,
  UK = NA
))

for(i in seq_along(fnam)) {
  load(file.path(root_dir, "Analysis", fnam[i]))
  results[[i]] <- list(
    data = x,
    model0 = x_gf,
    efp = x_efp,
    nbreaks = x_nbreaks,
    breakpoints = x_bp,
    model = x_rf,
    erm = erm[i],
    erm2 = erm2[i],
    eur = eur[i])
}
@

@@

\clearpage
This decrease in inflation is also conveyed by Figure~\ref{fig:slovenia2} that shows the
observed histograms of inflation rates before and after the breakpoint along with the
fitted GL distribution. This highlights that along with the decrease in mean level, the variance increases
somewhat and -- probably more interestingly -- the distribution changes from being right-skewed to
slightly left-skewed. This confirms that very high inflation rates that occurred in several
months before the breakpoint could be avoided afterwards. Furthermore, Figure~\ref{fig:slovenia2}
shows that the GL distribution fits the observed data very well, both before and after the
breakpoint, with $p$~values from $\chi^2$ goodness-of-fit tests of
\Sexpr{format(round(summary(x$model[[1]])$chisq.test$p.value, digits = 3), nsmall = 3)} and
\Sexpr{format(round(summary(x$model[[2]])$chisq.test$p.value, digits = 3), nsmall = 3)}, respectively.

@@

@@

<<eu, echo=FALSE>>=
## impose breaks "by hand"
eu_breaks <- yearmon(c(1998 + 11/12, 2008 + 7/12))
eu_seg <- 1 + (time(eu) > eu_breaks[1]) + (time(eu) > eu_breaks[2])
library("e1071")
eu_mom_old1 <- aggregate(eu, eu_seg, function(x) c(mean(x), var(x), skewness(x)))

## better: use model-implied moments
eu_mom_old2 <- aggregate(eu, eu_seg, function(x) moments(glogisfit(x)))

## or even better: estimate breaks?
## eu_bp <- breakpoints(glogisfit(eu), h = 24)
## moments(eu_bp, breaks = 2)
## replicate results "by hand"
eu_breaks <- yearmon(c(1999 + 5/12, 2007 + 7/12))
eu_seg <- 1 + (time(eu) > eu_breaks[1]) + (time(eu) > eu_breaks[2])
eu_mom <- aggregate(eu, eu_seg, function(x) moments(glogisfit(x)))
@

@@

Concerning inflation differentials, Figure~\ref{fig:glmean} depicts the inflation rates and a rolling mean estimation (with a window size of 12 months),
where the mean is calculated using the first moment of a fitted GL distribution.\footnote{If more countries have the same inflation rate in a given month,
the lines are overlapped and thus get darker. We wold see one black line if all countries would show the same month-on-month increase at any one time. The
red lines depict the means.} We do see that many of these countries follow a rather similar path. One event clearly visible is the financial crises and the
short deflationary phase starting somewhere around the autumn of 2007 and lasting a few months. The general trend also suggests, that the inflation rates
converged towards some common -- albeit broad -- mean level. For the inflation series of the whole EMU, as published by the \citet{oecd}, we see
an increase in mean from
$\Sexpr{round(eu_mom[1, 1], digits = 3)}$ (\Sexpr{start(eu)}--\Sexpr{eu_breaks[1]}) to
$\Sexpr{round(eu_mom[2, 1], digits = 3)}$ (\Sexpr{eu_breaks[1] + 1/12}--\Sexpr{eu_breaks[2]}) with a slight decrease to
$\Sexpr{round(eu_mom[3, 1], digits = 3)}$ (\Sexpr{eu_breaks[2] + 1/12}--\Sexpr{end(eu)}) following the financial crises.
Whereas variance has increased considerably over time from
$\Sexpr{round(eu_mom[1, 2], digits = 3)}$ over $\Sexpr{round(eu_mom[2, 2], digits = 3)}$ to $\Sexpr{round(eu_mom[3, 2], digits = 3)}$.
Regarding skewness, we witness a movement towards negative skewness from
$\Sexpr{round(eu_mom[1, 3], digits = 3)}$ over $\Sexpr{round(eu_mom[2, 3], digits = 3)}$ to $\Sexpr{round(eu_mom[3, 3], digits = 3)}$.

<<glmean>>=
if(!file.exists(file.path(root_dir, "Analysis", "glmean.rda"))) {
   glmean <- function(x) {
    if(any(is.na(x))) NA else glogisfit(x, hessian = FALSE)$moments[1]
  }  
  hicp_glmean <- rollapply(hicps, 12, glmean)
  setwd(file.path(root_dir, "Analysis"))
  save(hicp_glmean, file = "glmean.rda")
  setwd(file.path(root_dir, "Paper"))
} else {
  load(file.path(root_dir, "Analysis", "glmean.rda"))
}
@

>>>>

## additional moment method
moments <- function(object, ...) UseMethod("moments")
moments.glogisfit <- function(object, ...) object$moments
moments.breakpoints.glogisfit <- function(object, breaks = NULL, ...)
  t(sapply(refit(object, breaks = breaks), "[[", "moments"))

## convenience formatting function
round3 <- function(x) format(round(x, digits = 3), nsmall = 3)
@

@@

<<analysis-results>>=
## country names
nam <- c(colnames(hicps), "EU")

## run analysis (for seasonally adjusted data)
root_dir <- do.call("file.path", as.list(head(strsplit(getwd(), .Platform$file.sep)[[1]], -1)))
if(!all(file.exists(file.path(root_dir, "Analysis", paste(nam, "rda", sep = "."))))) {
setwd(file.path(root_dir, "Analysis"))
Sweave("results-adjusted.Rnw")
setwd(file.path(root_dir, "Paper"))
}

## set up named results list
results <- vector(length = length(nam), mode = "list")
names(results) <- nam

## Exchange Rate Mechanism information
## ERM~II, ERM~I (part of european monetary system expired in 1999-01-01 gave way to ERM II
## if never entered ""right?
## http://web.archive.org/web/20080101022356/http://lexikon.meyers.de/meyers/Europ%C3%A4isches_W%C3%A4hrungssystem
erm <- yearmon(c(
  Austria = 1995,
  Belgium = 1979 + 2/12,
  CzechRepubic = NA, 
  Denmark = 1979 + 2/12,
  Estonia = NA,
  Finland = 1996 + 9/12,
  France = 1979 + 2/12,
  Germany = 1979 + 2/12,
  Greece = 1998 + 2/12,
  Hungary = NA,
  Ireland = 1979 + 2/12,
  Italy = 1979 + 2/12,
  Luxembourg = 1979 + 2/12,
  Netherlands = 1979 + 2/12,
  Poland = NA,
  Portugal = 1992 + 3/12,
  Slovakia = NA,
  Slovenia = NA,
  Spain = 1986 + 5/12,
  Sweden = NA ,
  UK = NA,
  EU = NA
))

erm2 <- yearmon(c(
  Austria = NA,
  Belgium = NA,
  CzechRepubic = NA, 
  Denmark = 1999,
  Estonia = 2004 + 5/12,
  Finland = NA,
  France = NA,
  Germany = NA,
  Greece = 1999,
  Hungary = NA,
  Ireland = NA,
  Italy = NA,
  Luxembourg = NA,
  Netherlands = NA,
  Poland = NA,
  Portugal = NA,
  Slovakia = 2005 + 10/12,
  Slovenia = 2004 + 5/12,
  Spain = NA,
  Sweden = NA ,
  UK = NA,
  EU = NA
))

eur <- yearmon(c(
  Austria = 1999,
  Belgium = 1999,
  CzechRepubic = NA, 
  Denmark = NA,
  Estonia = 2011,
  Finland = 1999,
  France = 1999,
  Germany = 1999,
  Greece = 2001,
  Hungary = NA,
  Ireland = 1999,
  Italy = 1999,
  Luxembourg = 1999,
  Netherlands = 1999,
  Poland = NA,
  Portugal = 1999,
  Slovakia = 2009,
  Slovenia = 2007,
  Spain = 1999,
  Sweden = NA ,
  UK = NA,
  EU = NA
))

for(i in seq_along(nam)) {
  load(file.path(root_dir, "Analysis", paste(nam[i], "rda", sep = ".")))
  results[[i]] <- list(
    data = x,
    model0 = x_gf,
    efp = x_efp,
    nbreaks = x_nbreaks,
    breakpoints = x_bp,
    model = x_rf,
    erm = erm[i],
    erm2 = erm2[i],
    eur = eur[i])
}
@

@@

\clearpage
This decrease in inflation is also conveyed by Figure~\ref{fig:slovenia2} that shows the
observed histograms of inflation rates before and after the breakpoint along with the
fitted GL distribution. This highlights that along with the decrease in mean level, the variance increases
somewhat and -- probably more interestingly -- the distribution changes from being right-skewed to
slightly left-skewed. This confirms that very high inflation rates that occurred in several
months before the breakpoint could be avoided afterwards. Furthermore, Figure~\ref{fig:slovenia2}
shows that the GL distribution fits the observed data very well, both before and after the
breakpoint, with $p$~values from $\chi^2$ goodness-of-fit tests of
\Sexpr{round3(summary(x$model[[1]])$chisq.test$p.value)} and
\Sexpr{round3(summary(x$model[[2]])$chisq.test$p.value)}, respectively.

@@

\noalign{\medskip}
\multicolumn{8}{l}{\emph{European Union}} \\
\hline
<<moments-table6, results=tex>>=
writeLines(unlist(lapply(22, format_moments)))
@

@@

<<eu, echo=FALSE>>=
eu_breaks <- breakdates(results$EU$breakpoints, breaks = results$EU$nbreaks)
eu_mom <- moments(results$EU$breakpoints, breaks = results$EU$nbreaks)
@

@@

Concerning inflation differentials, Figure~\ref{fig:glmean} depicts the inflation rates and a rolling mean estimation (with a window size of 12 months),
where the mean is calculated using the first moment of a fitted GL distribution.\footnote{If more countries have the same inflation rate in a given month,
the lines are overlapped and thus get darker. We wold see one black line if all countries would show the same month-on-month increase at any one time. The
red lines depict the means.} We do see that many of these countries follow a rather similar path. One event clearly visible is the financial crises and the
short deflationary phase starting somewhere around the autumn of 2007 and lasting a few months. The general trend also suggests, that the inflation rates
converged towards some common -- albeit broad -- mean level. For the inflation series of the whole EMU, as published by the \citet{oecd}, we see
an increase in mean from
$\Sexpr{round3(eu_mom[1, 1])}$ (\Sexpr{start(eu)}--\Sexpr{eu_breaks[1]}) to
$\Sexpr{round3(eu_mom[2, 1])}$ (\Sexpr{eu_breaks[1] + 1/12}--\Sexpr{eu_breaks[2]}) with a slight decrease to
$\Sexpr{round3(eu_mom[3, 1])}$ (\Sexpr{eu_breaks[2] + 1/12}--\Sexpr{end(eu)}) following the financial crises.
Whereas variance has increased considerably over time from
$\Sexpr{round3(eu_mom[1, 2])}$ over $\Sexpr{round3(eu_mom[2, 2])}$ to $\Sexpr{round3(eu_mom[3, 2])}$.
Regarding skewness, we witness a movement towards negative skewness from
$\Sexpr{round3(eu_mom[1, 3])}$ over $\Sexpr{round3(eu_mom[2, 3])}$ to $\Sexpr{round3(eu_mom[3, 3])}$.

<<glmean>>=
glmean <- function(x) {
  if(any(is.na(x))) NA else glogisfit(x, hessian = FALSE)$moments[1]
}

if(!file.exists(file.path(root_dir, "Analysis", "glmean.rda"))) {
  hicp_glmean <- rollapply(hicps, 12, glmean)
  setwd(file.path(root_dir, "Analysis"))
  save(hicp_glmean, file = "glmean.rda")
  setwd(file.path(root_dir, "Paper"))
} else {
  load(file.path(root_dir, "Analysis", "glmean.rda"))
}
@

\fixme{need to note somewhere} Rapid changes that persist only for very short amounts of time
are not captured well. Specifically, short-lasting changes in the mean, may simply be attributed
to increased variance. For example, the mean inflation in the 
year before the crisis (\Sexpr{yearmon(2007 + 7/12)}--\Sexpr{yearmon(2008 + 6/12)}) was
$\Sexpr{round3(glmean(window(eu, start = yearmon(2007 + 7/12), end = yearmon(2008 + 6/12))))}$
and after the crisis (\Sexpr{yearmon(2008 + 7/12)}--\Sexpr{yearmon(2009 + 6/12)}) it dropped to
$\Sexpr{round3(glmean(window(eu, start = yearmon(2008 + 7/12), end = yearmon(2009 + 6/12))))}$
before rising again in the year after that (\Sexpr{yearmon(2009 + 7/12)}--\Sexpr{yearmon(2010 + 6/12)}):
$\Sexpr{round3(glmean(window(eu, start = yearmon(2009 + 7/12), end = yearmon(2010 + 6/12))))}$.
